---
title: "Introduction to meta-analysis"
author: "Steve Simon"
date: "11/14/2018"
output: html_document
---

```{r prelims, echo=FALSE}
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(metafor)))
opts_chunk$set(echo=FALSE, fig.width=6.5, fig.height=3.5) # was ht=4.5
bcg <- get(data(dat.bcg))
```

## Overview

This presentation will show how to use base R functions to reproduce the elements of a forest plot. It will also offer some suggestions on how you should interpret patterns in this plot.

## BCG description

"Results from 13 studies examining the effectiveness of the Bacillus Calmette-Guerin (BCG) vaccine against tuberculosis."

The BCG data comes with the metafor package in R, but is also available at https://rdrr.io/cran/metafor/man/dat.colditz1994.html

## BCG data, first three columns

```{r bcg-list1}
bcg[ , c("trial", "author", "year")]
```

## BCG data, last six columns

```{r bcg-list2}
bcg[ , c("tpos", "tneg", "cpos", "cneg", "ablat", "alloc")]
```

```{r bcg-fe, comment=""}
bcg_fixed_effects <- rma(
  method="FE",
  measure="OR", 
  ai=tpos, bi=tneg, 
  ci=cpos, di=cneg,
  data=bcg, slab=paste(author, year))
bcg_fixed_effects_text <-
  capture.output(bcg_fixed_effects)
cat(bcg_fixed_effects_text, sep="\n")
par(mar=c(3.6, 0, 0, 0))
forest(bcg_fixed_effects, cex=0.8)
```

Here are the results for the first study, listed in a two by two table. A positive result is listed in the first column and a negative result in the second column. A positive result is bad, indicating that the vaccine did not work.

```{r table}
v <- bcg[1, c("tpos", "tneg", "cpos", "cneg")]
m <- matrix(v, nrow=2, byrow=TRUE)
odds <- unlist(m[, 1])/unlist(m[, 2])
or <- odds[1]/odds[2]
log.or <- log(or)

tx <- function(d, y, txt, co) {
  text(d, y, txt, cex=0.8, adj=1, col=co)
}

display_table <- function(step, accent="red", i=1) {
  v <- bcg[i, c("tpos", "tneg", "cpos", "cneg")]
  w <- v # w <- format(v, big.mark=",")
  
  m <- matrix(v, nrow=2, byrow=TRUE)
  odds <- unlist(m[, 1])/unlist(m[, 2])
  or <- odds[1]/odds[2]
  log.or <- log(or)

  d0 <- -7.8
  d1 <- d0 + 0.9
  d2 <- d0 + 1.9
  d3 <- d0 + 5.2
  d4 <- d0 + 8.5
  d5 <- 4.75
  d6 <- d0 + 15.1

  y1 <- 15.8
  y2 <- y1 - 1.0
  y3 <- y1 - 1.5
  y4 <- 14.5 - i
  y5 <- 14.0 - i
  y6 <- 13.5 - i
  
  co <- ifelse(step==1, accent, "black")
  tx(d0, y1, "T", co)
  tx(d1, y1, w[1], co)
  tx(d2, y1, w[2], co)
  tx(d0, y2, "C", co)
  tx(d1, y2, w[3], co)
  tx(d2, y2, w[4], co)
  if (step==1) return(invisible(1))

  co <- ifelse(step==2, accent, "black")
  odds <- round(odds, 3)
  odds_t <- paste(w[1], "/", w[2], "=", odds[1])
  odds_c <- paste(w[3], "/", w[4], "=", odds[2])
  tx(d3, y1, odds_t, co)
  tx(d3, y2, odds_c, co)
  if (step==2) return(invisible(2))

  co <- ifelse(step==3, accent, "black")
  or <- round(or, 2)
  or_calc <- paste(odds[1], "/", odds[2], "=", or)
  tx(d4, y1, or_calc, co)
  if (step==3) return(invisible(3))
  
  co <- ifelse(step==4, accent, "black")
  log.or <- round(log.or, 2)
  log_calc <- paste0("ln(", or, ") = ", log.or)
  tx(d4, y2, log_calc, co)
  if (step==4) {
    segments(d4, y3, d4, y4, col=accent)
    segments(log.or, y4, 5.1, y4, col=accent)
    arrows(log.or, y4, log.or, y5, length=0.05, col=accent)
    segments(4.3, y6, 5.1, y6, col=accent)
    segments(4.3, y4, 4.3, y6, col=accent)
    segments(5.1, y4, 5.1, y6, col=accent)
  }
  if(step==4) return(invisible(4))

  co <- ifelse(step==5, accent, "black")
  var <- round(sum(1/v), 3)
  var_calc <- paste0(paste0("1/", v, collapse=" + "), " = ", var)
  tx(d6, y1, var_calc, co)
  if (step==5) return(invisible(5))

  co <- ifelse(step==6, accent, "black")
  lo <- round(log.or-1.96*sqrt(var), 2)
  hi <- round(log.or+1.96*sqrt(var), 2)
  ci_calc <- paste0(log.or, " ± 1.96 sqrt ", var," = ", lo, " to ", hi)
  tx(d6, y2, ci_calc, co)
  if (step==6) {
    segments(6.2, y3, 6.2, y4, col=accent)
    segments(7.1, y4, lo, y4, col=accent)
    arrows(lo, y4, lo, y5, length=0.05, col=accent)
    arrows(hi, y4, hi, y5, length=0.05, col=accent)
    segments(5.2, y6, 7.1, y6, col=accent)
    segments(5.2, y4, 5.2, y6, col=accent)
    segments(7.1, y4, 7.1, y6, col=accent)
  }
  if (step==6) return(invisible(6))
}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(1)
```

Calculate the odds for each row.

```{r odds, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(2)
```

The ratio of these two odds is the odds ratio.

```{r or, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(3)
```

The underlying distribution of the odds ratio is almost always skewed right and bounded below by zero, so you should consider a log transformation. This is the center of the confidence interval for the first study

```{r log-or, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4)
```

Repeat these calculations for the second,

```{r log-or2, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=2)
```

third,

```{r log-or3, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=3)
```

fourth,

```{r log-or4, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=4)
```

fifth,

```{r log-or5, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=5)
```

sixth,

```{r log-or6, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=6)
```

seventh,

```{r log-or7, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=7)
```

eighth,

```{r log-or8, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=8)
```

ninth,

```{r log-or9, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=9)
```

tenth,

```{r log-or10, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=10)
```

eleventh,

```{r log-or11, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=11)
```

twelvth,

```{r log-or12, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=12)
```

and thirteenth studies.

```{r log-or13, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(4, i=13)
```

The variance of the log odds ratio is approximately

$\frac{1}{n_{11}} + \frac{1}{n_{12}} + \frac{1}{n_{21}} + \frac{1}{n_{22}}$

where n11, n12, etc. are the count of each cell in the two by two table. For the first study, you get

```{r v, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(5)
```

The approximate confidence interval for the log odds is

```{r ci, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6)
```

Notice that this confidence interval includes the value of zero, so this particular study did not reject the null hypothesis.

Repeath this calculation for the second,

```{r ci2, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=2)
```

third,

```{r ci3, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=3)
```

fourth,

```{r ci4, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=4)
```

fifth,

```{r ci5, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=5)
```

sixth,

```{r ci6, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=6)
```

seventh,

```{r ci7, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=7)
```

eighth,

```{r ci8, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=8)
```

ninth,

```{r ci9, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=9)
```

tenth,

```{r ci10, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=10)
```

eleventh,

```{r ci11, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=11)
```

twelvth,

```{r ci12, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=12)
```

and thirteenth studies.

```{r ci13, echo=FALSE}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_table(6, i=13)
```

The log odds ratios and variances for the full data set are listed below.

```{r full-log-or, echo=FALSE}
bcg_or <- escalc(
  measure="OR", data=dat.bcg,
  ai=tpos, bi=tneg, 
  ci=cpos, di=cneg)
b <- data.frame(
  y=round(bcg_or$yi,2), 
  w=round(1/bcg_or$vi,1))
c <- b
names(c) <- c("log or", "var")
c
```

The overall estimate is a weighted average of the individual study estimates.

$\hat{\theta}=\frac{\Sigma \hat{\theta}_i w_i}{\Sigma w_i}$

```{r overall-weighted-sum}
theta_hat <- round(sum(b$y*b$w)/sum(b$w), 2)
l <- "("
r <- ")"
l %>%
  paste0(l) %>%
  paste0(b$y[1]) %>%
  paste0(r, l) %>%
  paste0(b$w[1]) %>%
  paste0(r, "+",l) %>%
  paste0(b$y[2]) %>%
  paste0(r, l) %>%
  paste0(b$w[2]) %>%
  paste0(r, "+...+",l) %>%
  paste0(b$y[13]) %>%
  paste0(r, l) %>%
  paste0(b$w[13]) %>%
  paste0(r, r) -> num_calc
l %>%
  paste0(b$w[1]) %>%
  paste0("+", b$w[2]) %>%
  paste0("+...+") %>%
  paste0(b$w[13]) %>%
  paste0(r) -> den_calc

d1 <- -4.9
d2 <- d1 + 3.3
d3 <- d2 + 0.8
d4 <- d3 + 0.5
d5 <- d4 + 0.7
d6 <- 7.3

y1 <- 15.8
y2 <- y1 - 0.5
y3 <- y2 - 0.5
y4 <- -0.5
y5 <- y4 - 0.5
y6 <- y5 - 0.5

display_overall <- function(step=1, accent="red") {
  co <- ifelse(step==1, accent, "black")
  text(d1, y1, num_calc, cex=0.8, adj=0.5, col=co)
  text(d1, y3, den_calc, cex=0.8, adj=0.5, col=co)
  segments(d1-3, y2, d1+3, y2, col=co)
  text(d2, y2, "=", cex=0.8, adj=0.5, col=co)
  text(d3, y1, round(sum(b$y*b$w), 1), cex=0.8, col=co)
  text(d3, y3, round(sum(b$w), 1), cex=0.8, col=co)
  segments(d3-0.4, y2, d3+0.4, y2, col=co)
  text(d4, y2, paste0("= ", theta_hat), cex=0.8, adj=0, col=co)
  if (step==1) {
    segments(d5, y3, d5, y4, col=co)
    segments(theta_hat, y4, 5.1, y4, col=co)
    segments(4.3, y6, 5.1, y6, col=co)
    segments(4.3, y4, 4.3, y6, col=co)
    segments(5.1, y4, 5.1, y6, col=co)
    arrows(theta_hat, y4, theta_hat, y5, length=0.05, col=co)
  }
  if (step==1) return(invisible(1))

  co <- ifelse(step==2, accent, "black")
  lo <- round(theta_hat - 1.96*sqrt(1/sum(b$w)), 2)
  hi <- round(theta_hat + 1.96*sqrt(1/sum(b$w)), 2)
  ci_calc <- paste0(theta_hat, " ± 1.96 sqrt (1/", round(sum(b$w), 1), ") = ", lo, " to ", hi)
  text(d6, y2, ci_calc, cex=0.8, adj=1, col=co)
  if (step==2) {
    segments(6.2, y3, 6.2, y4, col=accent)
    segments(7.1, y4, lo, y4, col=accent)
    arrows(lo, y4, lo, y5, length=0.05, col=accent)
    arrows(hi, y4, hi, y5, length=0.05, col=accent)
    segments(5.2, y6, 7.1, y6, col=accent)
    segments(5.2, y4, 5.2, y6, col=accent)
    segments(7.1, y4, 7.1, y6, col=accent)
  }
  return(invisible(2))
}
```


```{r overall_est}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_overall(1)
```

The variance of the overall estimate is simply the reciprocal of the sum of the weights. 

$V(\hat{\theta})=\frac{1}{\Sigma w_i}$

This is just the fraction seen above, but with the nuermator replaced by 1.

```{r overall_ci}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
display_overall(2)
```

The relative weights given to each study are listed below.

```{r wt}
f <- 0.2*sqrt(b$w/sum(b$w))
rw <- 100*b$w/sum(b$w)
rw <- ifelse(rw<=0.95, round(rw,1), round(rw))
x <- data.frame(
  study=c(1:13, "Total"),
  wt=c(b$w, sum(b$w)),
  rel_wt=paste0(c(rw, sum(rw)), "%"))
x
```

The relative weights determine the size of the squares in the forest plot.

```{r squares}
par(mar=c(3.6, 0, 0, 0)); forest(bcg_fixed_effects, cex=0.8)
text(b$y, (13:1)+0.5, x$rel_wt[-14], cex=0.5, col="red") 
```

Finally, notice a dotted vertical line at the log odds ratio=0. This is the null value.

## What to look for in a forest plot

The forest plot provides information about individual studies, and how they relate to the overall estimate. Here are some things to look for.

1. Look at the overall effect first. Does the overall confidence interval diamond intersect the vertical dotted line of no effect?

2. Count the number of individual studies which fail to achieve statistical significance. These are individual confidence intervals that intersect the vertical dotted line. The count of non-significant studies is not a primary measure of evidence, because you can easily have a meta-analysis where every single individual study is not statistically significant, but the overall result is strongly significant. Still, it's an interesting measure, and gives you an initial impression of what the exisiting publications have said so far.

3. Count the number of studies where the point estimate is on the "wrong" side of the dotted line. Again, this is not a primary measure of evidence. Informally, a meta-analysis strengthens the results of each individual study if all the individual studies are pointing in the same direction. A few studies pointing in the opposite direction will tend to dilute the strength of a meta-analytic finding (as they should).

4. Look at the relative weights given to each study. Does one study tend to dominate the findings?

5. Look for evidence of heterogeneity. This can occur several ways, but the easiest source of heterogeneity is a bimodal pattern in the studies. Most of the studies might show little or no effect, for example, but a smaller group of studies might cluster in a region of a strong effect. If you are familiar with the individual studies, ask yourself if there is a feature that is unique to the outlying cluster of studies.

### Funnel plot

```{r funnel}
b$se <- round(1/sqrt(b$w), 3)
theta_hat <- -0.44
par(mar=c(3.9, 3.9, 0, 0))
draw_funnel <- function() {
  funnel(bcg_fixed_effects)
  text(0.9, 0.04, expression(hat(theta[i])), adj=1, cex=0.8)
  text(1.3, 0.04, "Std.Err.", adj=1, cex=0.8)
  for (i in 1:13) {
    text(0.9, 0.04*(i+1), b$y[i], adj=1, cex=0.8)
    text(1.3, 0.04*(i+1), b$se[i], adj=1, cex=0.8)
  }
  text(0.9, 0.64, theta_hat, adj=1, cex=0.8)
  text(0.7, 0.64, "Overall", adj=1, cex=0.8)
  
}
draw_funnel()
arrows(0.7, 0.08, b$y[1], b$se[1], length=0.05, col="red")
d <- 0.365
arrows(-0.44-2*d, d, -0.44+2*d, d, length=0.05, col="red", code=3)
text(-0.44-2*d-0.02, d, paste0("+/-2*", d), adj=1, col="red", cex=0.8)
```

The center line is the overall estimate.

```{r save-everything, results=FALSE}
save.image(file="../data/calculations.RData")
```